\documentclass[11pt,a4paper]{article}
\usepackage[margin=2.4cm]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=blue!55!black,urlcolor=blue!55!black}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\uvec}{\mathbf{u}}
\newcommand{\zhat}{\hat{\mathbf{z}}}
\setlength{\parskip}{0.35em}
\setlength{\parindent}{0pt}

\title{\vspace{-1.5em}\textbf{A balance-consistent restart for the moving-cavity mesh change}\\[0.3em]
\large Deriving the conditions newly-opened and newly-iced cells must satisfy\\
for settle-free continuation \quad (rev.~2, post-review)}
\author{AWI-ESM3\,$\leftrightarrow$\,PISM moving cavity --- design note}
\date{2026-07-07}

\begin{document}
\maketitle

\begin{abstract}
\noindent
On every moving-cavity leg the FESOM restart is remapped onto a sub-mesh whose ice front has
moved. Empirically the leg then blows up within days: \emph{fast} as a vertical-CFL runaway at
$\Delta t=1200$\,s, or \emph{slow} as a barotropic $\eta_n\!\to\!$NaN at $\Delta t=60$\,s. We have
falsified every \emph{value-only} restart fix (removing static density inversions; nearest-neighbour
velocity; running without a settle). This note shows the two blow-ups are two faces of a single
\textbf{geostrophic adjustment}, that the balanced state a stable restart must be in is determined
by the pressure field (density $+$ ssh $+$ FESOM's cavity pressure boundary condition) together
with the surrounding spun-up water masses, and that the required velocity can therefore be
\textbf{computed} per level from $\nabla_h p$ --- one formula, no reference ambiguity. The plan
removes the \emph{leading-order} adjustment by construction; the residual discrete imbalance is
\emph{measured} before the run by explicit balance/stability metrics (\S\ref{sec:impl}, step~V).
\end{abstract}

\section{Status: what is falsified}\label{sec:status}
The instability is a coastal, mid-column ($\sim$95\,m) mode that grows over days from the changed
margin. Ruled out, each verified at the data level and each leaving the blow-up essentially
unchanged:
\begin{itemize}[leftmargin=1.5em,itemsep=1pt]
  \item \textbf{Static density inversions} --- a coherent-column $T/S$ refill removed all severe
        inversions; the leg still blew (\code{mstep}$\approx$164, CFLz$\approx$8.5).
        \emph{Metric (for reproducibility):} linearised EOS
        $\rho = 1027\,(1 - 1.7{\times}10^{-4}T + 7.6{\times}10^{-4}(S-35))$, adjacent-level
        $\Delta\rho$ over \emph{wet levels only} (\code{ulevels}\,..\,\code{nlevels}$-1$), minimum
        per column, over the 1595 retreated cells of the movcav8 chunk-3 sub-mesh: worst
        $-8.8\to-0.5$\,kg\,m$^{-3}$, cells with $\Delta\rho<-2$: $11\to0$. (Including dry
        sub-bottom padding or cavity levels above \code{ulevels} yields spurious $\sim{-27}$
        values from $S{=}0$/melt outliers --- do not compare against unmasked recomputes.)
  \item \textbf{Nearest-neighbour velocity} (Finn Heukamp / SO-ASE handling) --- filling changed
        elements with the nearest old element's current (no cell left at zero) lowered peak CFLz
        $8.5\to5.7$ but \emph{still} blew at \code{mstep}$\approx$164, now with a barotropic
        $\eta_n\!\to\!$NaN as well: the copied current is not balanced against the local density
        \emph{and} its depth-integral is not transport-consistent with $\eta$ (\S\ref{sec:transport}).
  \item \textbf{No settle / timestep} --- $\Delta t=60$\,s removed the fast CFLz mode entirely
        (0 warnings, $\sim$10 days) but a slow barotropic $\eta_n$ mode grew and NaN'd.
\end{itemize}
A value-only restart cannot be made \emph{balanced}: the imbalance is lateral (fronts) $+$ missing
momentum, not a vertical inversion.

\section{FESOM's cavity pressure convention (pinned from source)}\label{sec:convention}
Before writing balance conditions, the ssh/ice-load convention must be fixed; from
\code{oce\_ale\_pressure\_bv.F90} (\code{pressure\_force\_4\_linfs\_cavity} path, active for
\code{use\_cavity}):
\begin{itemize}[leftmargin=1.5em,itemsep=1pt]
  \item The ice-shelf load is \textbf{not} $\rho_i g h_i$. FESOM fills the cavity-occupied levels
        $1..\code{ulevels}{-}1$ with a \textbf{virtual water column carrying the ice-base $T,S$}
        (density re-evaluated at each level's pressure) and integrates it hydrostatically to get
        the pressure BC at the ice base. For a homogeneous ocean this construction is exactly
        pressure-gradient-free across the front (the code comment states this design goal).
  \item $\eta$ is \textbf{0 under the cavity} (verified in the evolved mother restart: all 1591
        sub-ice nodes carry $\eta=0.000$ exactly); the open-ocean $\eta$ is prognostic.
\end{itemize}
Hence the correct hydrostatic pressure at level $z$, everywhere, is
\begin{equation}
  p(z) \;=\; \rho_0\, g\, \eta \;+\; g\!\!\int_z^{0}\!\tilde\rho\,dz',\qquad
  \tilde\rho =
  \begin{cases}
    \rho(T,S,z') & \text{in the wet column},\\[2pt]
    \rho(T_b,S_b,z') & \text{in cavity-occupied levels } (T_b,S_b:\ \text{ice-base } T,S),
  \end{cases}
  \label{eq:pressure}
\end{equation}
with $\eta{=}0$ under ice. This replaces the $\rho_i g h_i$ form of rev.~1 (which was inconsistent
with the barotropic term there) and automatically handles the newly-iced margin: the ice load is
carried by the virtual column, so no separate ice-slope term is needed anywhere.

\section{The balance a stable restart must satisfy}\label{sec:balance}
"No adjustment shock" is equivalent to the restart satisfying, at every wet point:
\begin{align}
  \text{(1) static stability:}\quad & N^2 = -\tfrac{g}{\rho_0}\,\partial_z\rho \;\ge\; 0,\\
  \text{(2) hydrostatic pressure:}\quad & p(z)\ \text{per Eq.~\eqref{eq:pressure}},\\
  \text{(3) geostrophy, per level:}\quad & \uvec_g(z) = \tfrac{1}{f\rho_0}\,\zhat\times\nabla_h p(z),\\
  \text{(4) transport consistency:}\quad & \partial_t\eta = -\nabla_h\!\cdot\!\textstyle\int \uvec\,dz \approx 0,\\
  \text{(5) ALE consistency:}\quad & \textstyle\sum_k \code{hnode}_k = D + \eta \ \ (\text{no spurious } \partial_t h\to w).
\end{align}
Eq.~(3) is \textbf{one formula per level} --- thermal wind and the barotropic reference are both
contained in $\nabla_h p$, so there is no separate "integration constant" to choose and no
neighbour-velocity reference to add (rev.~1's step-3/step-4 split double-counted the barotropic
part; this supersedes it). (1) keeps convection off; (4) is what gates the \emph{slow} barotropic
$\eta_n$ blow-up and must be checked on the \emph{depth-integrated} field, not just pointwise; (5)
is a necessary condition that the \emph{current remap violates} (\code{remap\_hnode} assigns
nominal $z$-level thicknesses ignoring $\eta$) and therefore gets its own implementation step.

\emph{Continuous vs discrete:} geostrophic flow is non-divergent in the continuum; the entire
residual risk is the \emph{discrete} divergence at the changed front. The claim is therefore
"removes the leading-order adjustment", with the residual measured (step~V), not "balanced by
construction".

\section{Quantitative diagnosis: the missing jet}\label{sec:jet}
At the Weddell seed the fill leaves a front of $\Delta S\approx3$\,psu over $\Delta x\approx25$\,km:
$\Delta\rho\approx2.4$\,kg\,m$^{-3}$, $\nabla_h\rho\approx1.0\times10^{-4}$\,kg\,m$^{-4}$, and with
$f(-66^\circ)\approx-1.33\times10^{-4}$\,s$^{-1}$,
$\partial_z u = \tfrac{g}{f\rho_0}\nabla_h\rho \approx 7\times10^{-3}\,\text{s}^{-1}
\Rightarrow \Delta u \approx 0.7$\,m\,s$^{-1}$ over 100\,m.
\textbf{The balanced front carries a $\sim$0.7\,m\,s$^{-1}$ jet}; we initialised it at $0$ (v4) or
$\sim$0.06\,m\,s$^{-1}$ (v5) --- an $O(0.7)$ momentum deficit, which is the observed violent
adjustment. NN cannot supply it: the required current is specific to \emph{this} density gradient.

\section{Conditions per case}\label{sec:cases}
\textbf{Added points (ice retreated / newly wet).} At each newly-wet level: (i) $T,S$ laterally
\emph{anchored to the neighbouring water mass at that level} and vertically statically stable ---
the blend must be \textbf{density-monotonic} (enforce $N^2\ge0$ jointly on $T,S$, e.g.\ by
convective adjustment), \emph{not} independent $T,S$ interpolation, since mixing fresh ice-base
water below denser open-ocean water manufactures inversions; (ii) $\uvec$ from Eq.~(3);
(iii) $\eta$ from the open-ocean neighbours (0 if still sub-ice); (iv) \code{hnode} per (5).

\textbf{Points next to removed points (ice advanced / newly iced).} With the pinned convention
(\S\ref{sec:convention}) the new ice load is carried by the virtual-column pressure BC; the
adjacent open cells need no special term --- their Eq.~(3) velocity, computed from
Eq.~\eqref{eq:pressure}, already contains the geostrophic response to the new load.

\section{Corollary: do not over-sharpen the front}\label{sec:soften}
A $0.7$\,m\,s$^{-1}$ jet is high for an Antarctic cavity front, i.e.\ the current fill
over-sharpens it. \emph{Scope (corrected):} fully surface-opened columns already use a coherent
open-ocean donor; the "hold shelf-base water upward" path fires only for \textbf{still-sub-ice
thinned columns} --- which is exactly the observed seed (ulev $14\!\to\!2$, $S\approx31.0$ through
the column). The fix and its target are therefore: fill each newly-wet level of those columns from
the \textbf{nearest old node wet at that level} (lateral anchoring --- by construction the front
gradient matches the neighbouring unchanged cells, removing the free "softening knob"), then
enforce $N^2\ge0$ column-wise (density-monotonic adjustment). Gentler front $\Rightarrow$ smaller
required jet $\Rightarrow$ smaller residual after step~(3).

\section{Transport consistency (the slow mode)}\label{sec:transport}
$\partial_t\eta=-\nabla_h\!\cdot\!\int\uvec\,dz$: the slow $\eta_n$ blow-up is fed by any mismatch
between the depth-integrated velocity and $\eta$. Computing $\uvec$ per level from $\nabla_h p$
makes $\int\uvec\,dz$ automatically the geostrophic transport of the \emph{same} $p$ --- i.e.\
transport consistency holds to discretisation error. The verification therefore checks
$|\nabla_h\!\cdot(H\bar{\uvec})|$ explicitly (step~V), which is the quantity that gates the slow
mode; pointwise $|\nabla_h\!\cdot\uvec|$ alone is not sufficient.

\section{Implementation plan (\code{mo\_remap\_fields.F90} unless noted)}\label{sec:impl}
\begin{enumerate}[leftmargin=1.8em,itemsep=2pt,label=\textbf{\Roman*.}]
  \item \textbf{\code{hnode}/$\eta$ ALE consistency (new; cheapest, do first).} In
        \code{remap\_hnode}, scale the nominal thicknesses of changed columns so
        $\sum_k h_k = D + \eta$ (zstar distribution), with $\eta$ the staged (donor-filled, or 0
        sub-ice) value. Removes a standing $\partial_t h$/$\partial_t\eta$ source at $t{=}0$; a
        leading suspect for the slow mode.
  \item \textbf{Laterally-anchored, density-monotonic $T/S$ for still-sub-ice thinned columns.}
        Replace "hold shelf-base upward" by the per-level nearest-wet-old-node donor (memoised
        cache already exists), then a joint $T,S$ convective-adjustment pass over every changed
        column (also the donor-filled fully-opened ones) to guarantee $N^2\ge0$.
  \item \textbf{Geostrophic velocity on changed elements.} Post-pass over the written new-mesh
        restart: build $p(z)$ per node via Eq.~\eqref{eq:pressure} (linearised EOS; virtual
        column with ice-base $T,S$ under cavities; $+\rho_0 g\eta$, $\eta{=}0$ sub-ice); per-level
        $P_1$ gradient on each changed triangle; $\uvec_g=\tfrac{1}{f\rho_0}\zhat\times\nabla_h p$.
        Overwrite \code{u,v} there. No NN reference anywhere (supersedes v5; the NN fill remains
        only as a fallback where $|f|$ is degenerate, which does not occur on the Antarctic margin).
  \item \textbf{Zero the Adams--Bashforth history on changed elements}
        (\code{urhs\_AB,vrhs\_AB}): a balanced $\uvec$ with a foreign AB tendency reintroduces a
        1--2-step transient; zeroing gives a clean AB restart (FESOM handles first-step AB init).
  \item \textbf{Verify \emph{before} running} (offline, on the staged restart, changed cells):
        (a) $\min N^2$; (b) $\max|\nabla_h\!\cdot(H\bar{\uvec})|$ \emph{and} pointwise
        $|\nabla_h\!\cdot\uvec|$; (c) thermal-wind residual
        $|\partial_z\uvec-\tfrac{1}{f\rho_0}\zhat\times\nabla_h\partial_z p|$;
        (d) \textbf{stability of the constructed field}: advective CFL
        $|\uvec|\Delta t/\Delta x$, vertical shear Richardson number
        $Ri=N^2/|\partial_z\uvec|^2$ (flag $Ri<0.25$), and $\max|\uvec|$ sanity. The field must be
        both \emph{balanced} and \emph{integrable}.
  \item \textbf{Re-test} the movcav8 chunk-3 leg at $1200$\,s, no settle.
\end{enumerate}

\section{Caveats / open points}
\begin{itemize}[leftmargin=1.5em,itemsep=1pt]
  \item \textbf{Residual discrete divergence} at the changed front is the remaining risk; it is
        measured (step~V-b) rather than assumed away.
  \item \textbf{Ageostrophic background} (Ekman/tides) is not reconstructed; for a spun-up interior
        geostrophy dominates at $t{=}0$, and unchanged cells keep their full evolved velocity.
  \item \textbf{Forcing transient:} a balanced restart is steady only under steady forcing; the
        atmosphere will immediately flux the newly-opened cells. Settle-free means "no
        adjustment shock", not "forcing-free".
  \item \textbf{Still-sub-ice retreat cells} (the actual seed) get their $T/S$ from cavity-side
        per-level donors and $\eta=0$; their velocity still comes from Eq.~(3) --- the virtual
        column makes the pressure well-defined there.
  \item \textbf{Source term:} if the per-leg front motion is itself implausibly large (deep-draft
        transitions in one 10-yr PISM step), gentling the \code{cavity\_nlvls} regeneration
        reduces $\nabla_h\rho$ at the root and makes every step above easier.
\end{itemize}

\end{document}
